---
title: "Analysis"
output:
  html_document: default
  pdf_document: default
date: "2025-06-25"
authors: "Amanda Merino-Pelaz, Lucía Cobo-Sánchez"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

To open the data select file named dataset.

```{r data, , echo=FALSE}
library(FactoMineR)
library(factoextra)

x <- read.csv(file.choose(),header=T,sep=";") # dataset.csv
print(x)
```

#### Extract spatial parameters

```{r parameters, echo=FALSE}
params <- x[, c("Kappa","Scale","Clust_strength","Sibl_prob")]
print(params)
labels<-x[,1]
rownames(params)<-labels
```

## Dimensionality reduction: Principal Component Analysis (PCA)

Firstly, Bartlett's test of Sphericity and the Kaiser-Meyer-Olkin factor adequacy (KMO) were performed to assess the quality of the analyses.

#### Barlett test:
```{r barlett, echo=FALSE}
library("psych")
cortest.bartlett(params)#p.value <0.05
```

#### KMO test:
```{r kmo, echo=FALSE}
KMO(params)#Overall MSA >70
```

#### Multivariate outliers (Mahalanobis distance):
```{r outliers, echo=FALSE}
X <- scale(params)

# Distancia de Mahalanobis
md <- mahalanobis(X, 
                  center = colMeans(X), 
                  cov = cov(X))

# Umbral chi-cuadrado (97.5%)
threshold <- qchisq(0.975, df = ncol(X))

# Casos potencialmente influyentes
which(md > threshold)
```
#### Multivariate normality (Mardia):
```{r outliers, echo=FALSE}
library(MVN)

mvn(X)

```

#### PCA individual points (Figure 2a)
Individuals plot, colored by the quality of representation (cos²), with warmer colors indicating better representation in the component space.
```{r individuals, echo=FALSE}
res.pca <- PCA(params, scale.unit = TRUE, ncp = 5, graph = FALSE)
fviz_pca_ind(res.pca,
             col.ind = "cos2", # colored by quality of representation 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     
)
```

#### PCA variables (Figure 2b): 
Variables plot, colored by their contribution to the principal components, highlighting which metrics explain most of the variability in the dataset.
```{r PCAvariables, echo=FALSE}
fviz_pca_var(res.pca,
             col.var = "contrib", # colored by contribution
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE      
)
```

The PCA scores are then stored in the table under the name of scores_dim1 (PC1) and scores_dim2 (PC2).

```{r PCAscores, echo=FALSE}
ind_scores <- get_pca_ind(res.pca)

scores_dim1 <- ind_scores$coord[, 1]
scores_dim2 <- ind_scores$coord[, 2]

x$PC1 <- scores_dim1
x$PC2 <- scores_dim2
```

Finally, several PCA overlays are plotted. The potential associations observed with numeric variables are checked.

#### Figure 3a
Gradient-colored by chronological age (binned).

```{r PCAoverlay1, echo=FALSE}
library(ggplot2)
library(ggrepel)
my_palette <- colorRampPalette(c("darkblue","blue", "purple", "deeppink","orange", "yellow","green"))(100)

ggplot(x, aes(x = PC1, y = PC2, color = Chronology_ka)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = Site), size = 3, max.overlaps = 20) +
  scale_color_gradientn(colors = my_palette,
                        name = "Chronology (ka)",
                        breaks = seq(20, 120, by = 20)) +
  theme_minimal(base_size = 14) +
  labs(
    x = "PC1 (dispersion ↔ compactness)",
    y = "PC2 (cluster scale)"
  )
print("Spearman correlation between chronology and PC1 within UP:")
up <- subset(x, Period1 == "UP")
cor_up <- cor.test(up$Chronology, up$PC1, method = "spearman")
print(cor_up)

print("Spearman correlation between chronology and PC1 within MP:")
mp <- subset(x, Period1 == "MP")
cor_mp <- cor.test(mp$Chronology, mp$PC1, method = "spearman")
cor_mp
```

#### Figure 3b
Gradient-colored by mean intensity (m2).
```{r PCAoverlay2, echo=FALSE}
my_palette <- colorRampPalette(c("darkblue","blue", "purple", "deeppink","orange", "yellow","green"))(100)

ggplot(x, aes(x = PC1, y = PC2, color = Mean_intensity_m2)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = Site), size = 3, max.overlaps = 20) +
  scale_color_gradientn(colors = my_palette,
                        name = "Mean intensity",
                        breaks = seq(10, 500, by = 50)) +
  theme_minimal(base_size = 14) +
  labs(
    x = "PC1 (dispersion ↔ compactness)",
    y = "PC2 (cluster scale)"
  )

print("Spearman correlation between mean intensity and PC1:")
cor.test(x$PC1, x$Mean_intensity_m2, method = "spearman")

print("Spearman correlation between mean intensity and kappa:")
cor.test(x$Kappa, x$Mean_intensity_m2, method = "spearman")

print("Spearman correlation between mean intensity and sibling probability:")
cor.test(x$Sibl_prob, x$Mean_intensity_m2, method = "spearman")

print("Spearman correlation between mean intensity and cluster strength:")
cor.test(x$Clust_strength, x$Mean_intensity_m2, method = "spearman")

print("Spearman correlation between mean intensity and scale:")
cor.test(x$Scale, x$Mean_intensity_m2, method = "spearman")

```


#### Figure 3c
Colored by period (Middle Paleolithic, Upper Paleolithic) and shaped by site context (open-air, cave). Ellipses represent 95% confidence for each group OA-MP, OA-UP, C-MP, C-UP.
```{r PCAoverlay3, echo=FALSE}
ggplot(x, aes(x = PC1, y = PC2, color = Period1, shape = Site_context)) +
  geom_point(size = 3) +
  stat_ellipse(level = 0.95, linetype = "dashed") +
  theme_minimal(base_size = 14) +
  labs(
    x = "PC1 (dispersion ↔ compactness)",
    y = "PC2 (Cluster Scale)",
    color = "Period",
    shape = "Site context"
  )

```


#### Figure 3d 
Colored by sub-period and shaped by site context (open-air, cave). Ellipses represent 95% confidence for periods MP and UP.
```{r PCAoverlay4, echo=FALSE}
ggplot(x, aes(x = PC1, y = PC2, color = Period2, shape = Site_context)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = Period1, linetype = Period1), level = 0.95, show.legend = FALSE) +
  scale_color_brewer(palette = "Set2") +
  scale_shape_manual(values = c(16, 17)) +
  theme_minimal(base_size = 14) +
  labs(
    x      = "PC1 (dispersion ↔ compactness)",
    y      = "PC2 (cluster scale)",
    color  = "Sub-period",
    shape  = "Site context"
  )
```

#### Figure 3e
Colored by functionality of site (residential and non-residential) and shaped by length of occupation (short-term, long-term). Ellipses represent 95% confidence for residential and non-residential groups.  
```{r PCAoverlay5, echo=FALSE}
ggplot(x, aes(x = PC1, y = PC2, color = Functionality_bi, shape = Duration)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = Functionality_bi, linetype = Functionality_bi), level = 0.95, show.legend = FALSE) +
  scale_color_brewer(palette = "Set2") +
  scale_shape_manual(values = c(16, 17)) +
  theme_minimal(base_size = 14) +
  labs(
    x      = "PC1 (dispersion ↔ compactness)",
    y      = "PC2 (cluster scale)",
    color  = "Functionality",
    shape  = "Length of occupation"
  )
```


## Variable screening and transformations

Prior to the regressions, the distribution of  numeric and categorical variables (functionality, duration, site context, period, chronology, number of hearths, area, and mean intensity) were explored by means of histograms, Q-Q plots, Shapiro-Wilk test, and pairwise correlations (Pearson/Spearman correlations, variance inflation factors). Detailed results of this process are in the suplementary of materials sections IV (categorical) and V (numeric).

### Period

#### SOM section IV Figure 2a 
Violin, box and jitter plots showing the distribution of PC1 scores grouped by period (Middle Paleolithic vs. Upper Paleolithic).
```{r periodfig, echo=FALSE}
ggplot(x, aes(x = Period1, y = PC1, fill = Period1)) +
  # 1) Violin para la densidad
  geom_violin(trim = FALSE, alpha = 0.5) +
  # 2) Boxplot encima, estrecho para no tapar el violín
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.8) +
  # 3) Jitter de puntos individuales
  geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  # Estética
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal(base_size = 14) +
  labs(
    x = "Period",
    y = "PC1 (dispersion ↔ compactness)"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 12),
    axis.title = element_text(size = 13)
  )
```

#### Descriptive statistics and tests:

```{r periodstats, echo=FALSE}
print("Descriptive statistics:")
library(rstatix)
library(dplyr)
x %>% group_by(Period1) %>%
  summarise(
    n = n(),
    mean = mean(PC1),
    sd   = sd(PC1),
    median = median(PC1),
    IQR = IQR(PC1)
  )

print("Welch two‐sample t-test:")
t.test(PC1 ~ Period1, data=x, var.equal=FALSE)

print("Cohen's d:")
cohen.d(PC1 ~ Period1, data = x)

print("Bootstrap CI of the difference between means:")

library(boot)
set.seed(123)
boot_diff <- function(data, i) {
  d <- data[i, ]
  mean(d$PC1[d$Period1=="UP"]) - mean(d$PC1[d$Period1=="MP"])
}
b <- boot(x, boot_diff, R = 2000)
boot.ci(b, type = "perc")
 
print("Shapiro test applied to each group:")

mp <- filter(x, Period1=="MP")
up <- filter(x, Period1=="UP")

print("MP:")
shapiro_mp <- shapiro.test(mp$PC1)
shapiro_mp

print("UP:")
shapiro_up <- shapiro.test(up$PC1)
shapiro_up

print("Wilcoxon rank-sum test:")
wilcox.test(PC1 ~ Period1, data=x)
```


### Site context

#### SOM section IV Figure 2b
Violin, box and jitter plots showing the distribution of PC1 scores grouped by site context (open-air vs. cave).
```{r contextfig, echo=FALSE}

ggplot(x, aes(x = Site_context, y = PC1, fill = Site_context)) +
  # 1) Violin para la densidad
  geom_violin(trim = FALSE, alpha = 0.5) +
  # 2) Boxplot encima, estrecho para no tapar el violín
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.8) +
  # 3) Jitter de puntos individuales
  geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  # Estética
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal(base_size = 14) +
  labs(
    x = "Site Context",
    y = "PC1 (dispersion ↔ compactness)"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 12),
    axis.title = element_text(size = 13)
  )
```

#### Descriptive statistics and tests:

```{r contextstats, echo=FALSE}
print("Descriptive statistics:")
x %>% group_by(Site_context) %>%
  summarise(
    n = n(),
    mean = mean(PC1),
    sd   = sd(PC1),
    median = median(PC1),
    IQR = IQR(PC1)
  )

print("Levene's test:")
car::leveneTest(PC1 ~ Site_context, data=x)
  
print("Welch two‐sample t-test:")
t.test(PC1 ~ Site_context, data=x, var.equal=FALSE)

print("Wilcoxon rank-sum test:")
wilcox.test(PC1 ~ Site_context, data=x)
```


### Functionality

#### SOM section IV Figure 2c 
Violin, box and jitter plots showing the distribution of PC1 scores grouped by site functionality (residential vs. non-residential).
```{r functfig, echo=FALSE}

ggplot(x, aes(x = Functionality_bi, y = PC1, fill = Functionality_bi)) +
  # 1) Violin para la densidad
  geom_violin(trim = FALSE, alpha = 0.5) +
  # 2) Boxplot encima, estrecho para no tapar el violín
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.8) +
  # 3) Jitter de puntos individuales
  geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  # Estética
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal(base_size = 14) +
  labs(
    x = "Functionality",
    y = "PC1 (dispersion ↔ compactness)"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 12),
    axis.title = element_text(size = 13)
  )
```

#### Descriptive statistics and tests:

```{r functstats, echo=FALSE}
print("Descriptive statistics:")
x %>% group_by(Functionality_bi) %>%
  summarise(
    n = n(),
    mean = mean(PC1),
    sd   = sd(PC1),
    median = median(PC1),
    IQR = IQR(PC1)
  )

print("Levene's test:")
car::leveneTest(PC1 ~ Functionality_bi, data=x)
  
print("Welch two‐sample t-test:")
t.test(PC1 ~ Functionality_bi, data=x, var.equal=FALSE)

print("Wilcoxon rank-sum test:")
wilcox.test(PC1 ~ Functionality_bi, data=x)
```

### Chronology
#### SOM section V Figure 3a 
Density histogram of chronology in which the y-axis represents probability density, blue line represents kernel density estimates to visualize the underlying distribution of the data
```{r chronfig1, echo=FALSE}
ggplot(x, aes(x = Chronology_ka)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 10, color = "black", fill = "white") +
  geom_density(color = "blue", size = 1) +
  labs(
       x = "Chronology",
       y = "Density") +
  theme_minimal()
```

#### SOM section V Figure 4a
Q–Q plot of chronology comparing the distribution of the data with a theoretical normal distribution.
```{r chrontfig2, echo=FALSE}
ggplot(x, aes(sample = Chronology_ka)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Chronology") +
  theme_minimal()
```

#### Shapiro-Wilk test for normality:
```{r chronstats, echo=FALSE}
shapiro.test(x$Chronology_ka)
```


### Number of hearths
#### SOM section V Figure 3b
Density histogram of hearths count in which the y-axis represents probability density, blue line represents kernel density estimates to visualize the underlying distribution of the data
```{r hearthsfig1, echo=FALSE}
ggplot(x, aes(x = Num_hearths)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 10, color = "black", fill = "white") +
  geom_density(color = "blue", size = 1) +
  labs(
       x = "Number of hearths",
       y = "Density") +
  theme_minimal()
```

#### SOM section V Figure 4b 
Q–Q plot of hearths count comparing the distribution of the data with a theoretical normal distribution.
```{r hearthsfig2, echo=FALSE}
ggplot(x, aes(sample = Num_hearths)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Number of Hearths") +
  theme_minimal()
```

#### Shapiro-Wilk test for normality:
```{r hearthsstats1, echo=FALSE}
shapiro.test(x$Num_hearths)
```

#### Correlation tests with clustering metrics:
```{r hearthsstats2, echo=FALSE}
cor.test(x$Kappa, x$Num_hearths, method = "spearman")
cor.test(x$Sibl_prob, x$Num_hearths, method = "spearman")
cor.test(x$Clust_strength, x$Num_hearths, method = "spearman")
cor.test(x$Scale, x$Num_hearths, method = "spearman")
```

Log transformation of number of hearths.
```{r periodstats3, echo=FALSE}
x <- x%>%mutate(log_hearths = log(Num_hearths + 1))
```


### Excavated area
#### SOM section V Figure 3c 
Density histogram of area in which the y-axis represents probability density, blue line represents kernel density estimates to visualize the underlying distribution of the data
```{r areafig1, echo=FALSE}
ggplot(x, aes(x = Excavated_area_m2)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 10, color = "black", fill = "white") +
  geom_density(color = "blue", size = 1) +
  labs(
       x = "Area",
       y = "Density") +
  theme_minimal()
```

#### SOM section V Figure 4c
Q–Q plot of area comparing the distribution of the data with a theoretical normal distribution.
```{r areafig2, echo=FALSE}
ggplot(x, aes(sample = Excavated_area_m2)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Area") +
  theme_minimal()
```

#### Shapiro-Wilk test for normality:
```{r areastats1, echo=FALSE}
shapiro.test(x$Excavated_area_m2)
```

#### Correlation tests with clustering metrics:
```{r areastats2, echo=FALSE}
cor.test(x$Kappa, x$Excavated_area_m2, method = "spearman")
cor.test(x$Sibl_prob, x$Excavated_area_m2, method = "spearman")
cor.test(x$Clust_strength, x$Excavated_area_m2, method = "spearman")
cor.test(x$Scale, x$Excavated_area_m2, method = "spearman")
```

#### Variation of area among periods;
```{r areastats3, echo=FALSE}
print("Levene's test:")
car::leveneTest(Excavated_area_m2 ~ Period1, data=x)

print("Levene's test in open-air sites:")
x_open <- subset(x, Site_context == "OA")
car::leveneTest(Excavated_area_m2 ~ Period1, data = x_open)

print("Levene's test in cave sites:")
x_cave <- subset(x, Site_context == "C")
car::leveneTest(Excavated_area_m2 ~ Period1, data = x_cave)
```


### Mean intensity
#### SOM section V Figure 3d
Density histogram of mean intensity in which the y-axis represents probability density, blue line represents kernel density estimates to visualize the underlying distribution of the data
```{r intfig1, echo=FALSE}
ggplot(x, aes(x = Mean_intensity_m2)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 10, color = "black", fill = "white") +
  geom_density(color = "blue", size = 1) +
  labs(
       x = "Mean intensity",
       y = "Density") +
  theme_minimal()
```

#### SOM section V Figure 4d
Q–Q plot of mean intensity comparing the distribution of the data with a theoretical normal distribution.
```{r intfig2, echo=FALSE}
ggplot(x, aes(sample = Mean_intensity_m2)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Mean intensity") +
  theme_minimal()
```

#### Shapiro-Wilk test for normality:
```{r intstats1, echo=FALSE}
shapiro.test(x$Mean_intensity_m2)
```

#### Correlation tests with clustering metrics:
```{r intstats2, echo=FALSE}
cor.test(x$Kappa, x$Mean_intensity_m2, method = "spearman")
cor.test(x$Sibl_prob, x$Mean_intensity_m2, method = "spearman")
cor.test(x$Clust_strength, x$Mean_intensity_m2, method = "spearman")
cor.test(x$Scale, x$Mean_intensity_m2, method = "spearman")
```

## Regression models

Models were fitted with variables period, context, functionality, and number of hearths.

### Models with PC1

#### Full OLS model (Model 1, n=22) and multicollinearity test (Variance Inflation Factors):
```{r model1, echo=FALSE}
m_PC1 <- lm(PC1 ~ Period1 + Site_context + log_hearths + Functionality_bi, data = x)
summary(m_PC1)
m_PC1 <- lm(PC1 ~ Period1 + Site_context + log_hearths + Chronology_ka + Functionality_bi + Duration, data = x)

library(car)
vif(m_PC1)
```
##### Model validation
```{r modelval1, echo=FALSE}
print("1. Normality of residuals:")
qqnorm(residuals(m_PC1)); qqline(residuals(m_PC1))
shapiro.test(residuals(m_PC1))

print("2. Homocedasticity:")
plot(fitted(m_PC1), residuals(m_PC1))
abline(h=0, lty=2)
#Test de Breusch–Pagan:
library(lmtest)
bptest(m_PC1)

print("3. Independency of residuals:")
dwtest(m_PC1)

print("4. Influence/leverage. Cook's disance:")
plot(cooks.distance(m_PC1), type="h")

thr <- 4/(nrow(x) - length(coef(m_PC1)))
infl_points <- which(cooks.distance(m_PC1) > thr)
infl_points

```

#### Full OLS model without El Salt (Model 2, n=21):
```{r model2, echo=FALSE}
m_PC1_nolim <- lm(PC1 ~ Period1 + Site_context + log_hearths + Functionality_bi,data = x[-infl_points, ])
summary(m_PC1_nolim)

print("AIC Model 1 vs Model 2:")
AIC(m_PC1,m_PC1_nolim)
```

##### Model validation
```{r modelval2, echo=FALSE}
print("1. Normality of residuals:")
qqnorm(residuals(m_PC1_nolim)); qqline(residuals(m_PC1_nolim))
shapiro.test(residuals(m_PC1_nolim))

print("2. Homocedasticity:")
plot(fitted(m_PC1_nolim), residuals(m_PC1_nolim))
abline(h=0, lty=2)
#Test de Breusch–Pagan:
bptest(m_PC1_nolim)

print("3. Independency of residuals:")
dwtest(m_PC1_nolim)

print("4. Influence/leverage. Cook's disance:")
plot(cooks.distance(m_PC1_nolim), type="h")
which(cooks.distance(m_PC1_nolim) > 4/(nrow(x)-length(coef(m_PC1_nolim))))
```


#### Parsimonious OLS model (Model 3, n=22):
```{r model3, echo=FALSE}
m_simple <- lm(PC1 ~ Period1, data = x)
summary(m_simple)

print("AIC Model 1 vs Model 3:")
AIC(m_simple, m_PC1)
```
##### Model validation
```{r modelval3, echo=FALSE}
print("1. Normality of residuals:")
qqnorm(residuals(m_simple)); qqline(residuals(m_simple))
shapiro.test(residuals(m_simple))

print("2. Homocedasticity:")
plot(fitted(m_simple), residuals(m_simple))
abline(h=0, lty=2)
#Test de Breusch–Pagan:
bptest(m_simple)

print("3. Independency of residuals:")
dwtest(m_simple)
```


#### Robust OLS model (Model 4, n=22):
```{r model4, echo=FALSE}
library(MASS)
m_robust <- rlm(PC1 ~ Period1 + Site_context + log_hearths + Functionality_bi, data = x)
summary(m_robust)
```
##### Model validation

```{r modelval4, echo=FALSE}
print("1. Normality of residuals:")
qqnorm(residuals(m_robust)); qqline(residuals(m_robust))
shapiro.test(residuals(m_robust))

print("2. Homocedasticity:")
plot(fitted(m_robust), residuals(m_robust))
abline(h=0, lty=2)
#Test de Breusch–Pagan:
bptest(m_robust)

print("3. Independency of residuals:")
dwtest(m_robust)
```

#### Checking the robustness of period coefficient

```{r robustness, echo=FALSE}
boot_coef <- function(data, indices) {
  d <- data[indices, ]
  return(coef(lm(PC1 ~ Period1, data = d))["PeriodUP"])
}

set.seed(123)   
bstrap <- boot(
  data      = x,
  statistic = boot_coef,
  R         = 2000
)

print(bstrap)


boot_ci <- boot.ci(bstrap, type = "perc")
print(boot_ci)

hist(bstrap$t, 
     breaks = 30, 
     main   = "Bootstrap distribution of PeriodUP coef",
     xlab   = "Coefficient value",
     border = "white")
abline(v = quantile(bstrap$t, c(0.025, 0.975)), col = "red", lwd = 2, lty = 2)
abline(v = bstrap$t0, col = "blue", lwd = 2)
```

### Model with PC2

#### Full OLS model (n=22):
```{r PC2model1, echo=FALSE}
m_PC2 <- lm(PC2 ~ Period1 + Site_context + log_hearths + Functionality_bi, data = x)
summary(m_PC2)
```
##### Model validation
```{r PC2modelval1, echo=FALSE}
print("1. Normality of residuals:")
qqnorm(residuals(m_PC2)); qqline(residuals(m_PC2))
shapiro.test(residuals(m_PC2))

print("2. Homocedasticity:")
plot(fitted(m_PC2), residuals(m_PC2))
abline(h=0, lty=2)
#Test de Breusch–Pagan:
library(lmtest)
bptest(m_PC2)

print("3. Independency of residuals:")
dwtest(m_PC2)

print("4. Influence/leverage. Cook's disance:")
plot(cooks.distance(m_PC2), type="h")
which(cooks.distance(m_PC2) > 4/(nrow(x)-length(coef(m_PC2))))#Señala los casos influyentes.
hatvalues(m_PC2)[19]
hatvalues(m_PC2)[13]
```

#### Full OLS model without El Salt and Molodova (n=20):
```{r PC2model2, echo=FALSE}
m_PC2_nolim <- lm(PC2 ~ Period1 + Site_context + log_hearths + Functionality_bi,data = x[-13,-19, ])
summary(m_PC2_nolim)

AIC(m_PC2, m_PC2_nolim)
```

### Sensitivity analysis
```{r PC2model2, echo=FALSE}
print("Exploring individual metrics effects")
m_cs <- lm(Clust_strength ~ Period1 + Site_context + log_hearths + Functionality_bi, data = x)
summary(m_cs)

m_sp <- lm(Sibl_prob ~ Period1 + Site_context + log_hearths + Functionality_bi, data = x)
summary(m_sp)

m_k <- lm(Kappa ~ Period1 + Site_context + log_hearths + Functionality_bi, data = x)
summary(m_k)

m_s <- lm(Scale ~ Period1 + Site_context + log_hearths + Functionality_bi, data = x)
summary(m_s)


print("Controling excavated area effect:")

m_sens <- lm(PC1 ~ Period1 + Site_context + log_hearths + Functionality_bi + Excavated_area_m2, data = x)
summary(m_sens)

m_sens2 <- lm(Clust_strength ~ Period1 + Site_context + log_hearths + Functionality_bi + Excavated_area_m2, data = x)
summary(m_sens2)

m_sens3 <- lm(Sibl_prob ~ Period1 + Site_context + log_hearths + Functionality_bi + Excavated_area_m2, data = x)
summary(m_sens3)

m_sens4 <- lm(Kappa ~ Period1 + Site_context + log_hearths + Functionality_bi + Excavated_area_m2, data = x)
summary(m_sens4)

m_sens5 <- lm(Scale ~ Period1 + Site_context + log_hearths + Functionality_bi + Excavated_area_m2, data = x)
summary(m_sens5)


print("Controling compactness index effect:")
m_sens6 <- lm(PC1 ~ Period1 + Site_context + log_hearths + Functionality_bi+ Compactness, data = x)
summary(m_sens6)

m_sens7 <- lm(Clust_strength ~ Period1 + Site_context + log_hearths + Functionality_bi + Compactness, data = x)
summary(m_sens7)

m_sens8 <- lm(Sibl_prob ~ Period1 + Site_context + log_hearths + Functionality_bi + Compactness, data = x)
summary(m_sens8)

m_sens9 <- lm(Kappa ~ Period1 + Site_context + log_hearths + Functionality_bi + Compactness, data = x)
summary(m_sens9)

m_sens10 <- lm(Scale ~ Period1 + Site_context + log_hearths + Functionality_bi + Compactness, data = x)
summary(m_sens10)

print("Controling mean intensity effect:")
m_sens11 <- lm(PC1 ~ Period1 + Site_context + log_hearths + Functionality_bi+ Mean_intensity_m2, data = x)
summary(m_sens11)

m_sens12 <- lm(Clust_strength ~ Period1 + Site_context + log_hearths + Functionality_bi + Mean_intensity_m2, data = x)
summary(m_sens12)

m_sens13 <- lm(Sibl_prob ~ Period1 + Site_context + log_hearths + Functionality_bi + Mean_intensity_m2, data = x)
summary(m_sens13)

m_sens14 <- lm(Kappa ~ Period1 + Site_context + log_hearths + Functionality_bi + Mean_intensity_m2, data = x)
summary(m_sens14)

m_sens15 <- lm(Scale ~ Period1 + Site_context + log_hearths + Functionality_bi + Mean_intensity_m2, data = x)
summary(m_sens15)

```


### Figure 4
Chronological sequence of clustering metrics. A) PC1 scores, B) Cluster strength. C) Parent intensity kappa.  

```{r fig4, echo=FALSE}
ggplot(x, aes(x = Chronology_ka, y = PC1, color = Period1)) +
  geom_point(size = 3) +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", color = "grey40") +
  geom_text_repel(aes(label = Site), size = 3, color = "black", max.overlaps = Inf) +
  scale_x_reverse() +
  labs(x = "Chronology (ka)", y = "PC1") +
  theme_minimal()

ggplot(x, aes(x = Chronology_ka, y =Clust_strength, color = Period1)) +
  geom_point(size = 3) +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", color = "grey40") +
  geom_text_repel(aes(label = Site), size = 3, color = "black", max.overlaps = Inf) +
  scale_x_reverse() +
  labs(x = "Chronology (ka)", y = "Cluster strength") +
  theme_minimal()

ggplot(x, aes(x = Chronology_ka, y =Kappa, color = Period1)) +
  geom_point(size = 3) +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", color = "grey40") +
  geom_text_repel(aes(label = Site), size = 3, color = "black", max.overlaps = Inf) +
  scale_x_reverse() +
  labs(x = "Chronology (ka)", y = "Parent intensity") +
  theme_minimal()

```
